---
title: "dplyr_notes"
output: html_document
date: '2022-07-13'
---
# Data Processing <br>
### Introduction to `dplyr`

DPLYR is an R package in the tidyverse. It stands for, none other than, **data plier**! Just a fun way to spell it. We can load the package using ...

```{r}
##install.packages("dplyr")
library(dplyr)
library(ggplot2)

```

We can use dplyr to manipulate the data so that missing information in the data has the proper value and not some random value that was in the original set. You can change data as well.

```{r}
#loading required libraries for practice
head(starwars)

```

There are many main functions...

### Filtering Data <br>
We can filter data using `filter()`. It allows us to subset based on the data values themselves.

* Be sure you spell the column name correctly (and the value name if its a categorial variable). Remember, R is case sensitive
* Be sure yo use `==` when comparing observations. (Remember `=` is an assignment operator).
* You can use >, <, >=, <= to compare numeric or categorical variables (nominal variables are ranked alphabetically, while ordinal variables have a built-in rank).

```{r}
filter(starwars, hair_color == "blond" | hair_color == "blonde")
#uses expressions that does the filtering, double equals because its a comparison
#single bar '|' for or, & for and
#you can add as many values as you want

#you can do multiple values with commas
filter(starwars, species == "Human", homeworld == "Tatooine", skin_color == "fair")

#to store it as a variable
filteredData <- filter(starwars, addVariablesHere)
                       
filter(starwars, hair_color %in% c("blond","blonde"))


importantPlanets <- c("Tatooine", "Alderaan", "Naboo")
filter(starwars, homeworld %in% importantPlanets)
```

## Ranking Data <br> <br>
descr
```{r}
head(diamonds) #first six rows
class(diamonds$cut) #says the types
#its an ordered factor
summary(diamonds$cut) #shows count of each category or sumarry statistics (is numeric)
head(diamonds$cut) #first six values of that row (and levels, showing whats better than what)

filter(diamonds, cut > "Good")

#ordering categorical data
unique(starwars$eye_color)
factor(starwars$eye_color, c("red", "red, blue", "pink", "orange", "brown", "hazel", "gold", "yellow", "green, yellow", "blue", "blue-gray", "dark", "black", "white", "unknown"), ordered = T)
#^changes it from nominal to ordinal data

#practice
#find all characters that are shorter than 100cm
filter(starwars, height < 100, species != "Droid")

#which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year >= 100 & birth_year <= 200)
#shortcut!!
filter(starwars, between(birth_year, 100, 200))

#which character weigh over 100kg but are shorter than 185cm
filter(starwars, mass > 100 & height < 185)

```

## Missing Values
```{r}
NA > 5 #it doesnt know

mass <- NA
mass == 10 #returns NA
is.na(NA) #returns true


#which characters are missinga hair color
filter(starwars,is.na(hair_color))
```


## Arranging Data
`arrange()` reorders rows. It does not remove any rows.
NA values are always at the end when you order by a column.
```{r}
#lowest to highest birth year
arrange(starwars, birth_year)


#highest to lowest birth year
arrange(starwars, desc(birth_year))

arrange(starwars, hair_color) #will be alphabetically

#multiple columns
arrange(starwars, hair_color, birth_year)


## practice!
#arrange characters to find the tallest characters and the shortest characters
arrange(starwars, height) #finding shortest person
arrange(starwars, desc(height)) #finding tallest person


#alphabetize the characters
arrange(starwars, name)

#how could you use arrange() to sort all missing values to the start?
arrange(starwars, desc(is.na(starwars)))


```

## Renaming Variables
Write the new name first, then the original name
```{r}
rename(starwars, birthYear = birth_year)


#practice
```



## Select
We can select certain columns in the dataset

`select()` allows us to retain only certain variables(columns). It doesnt change the order, but it  removes other columns.

This is also a way to anonymize data, showing only certain elemens

```{r}
select(starwars, hair_color, skin_color, eye_color)

select(starwars, hair_color:eye_color) #returns in betweens (inclusive)

select(starwars, -hair_color)
select(starwars, -(hair_color:eye_color))

#wanting the columns that have the word color in them
select(starwars, contains("color"))

#we can use select() to reorder columns
#saying which columns you want first, then the rest
#you could also just list them all out
select(starwars, species, everything())


#practice
#select out the homeworld and species for the starwars dataset
select(starwars, homeworld, species)


#brainstorm as many possible ways to select hair_color, eye_color, skin_color, and birth_year
select(starwars, -(name:mass),  -(sex:starships))
select(starwars, hair_color, eye_color, skin_color, birth_year)
select(starwars, contains("_"))
select(starwars, ends_with("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars, ends_with("color"), ends_with("year"))
#so many options...

```


## Adding new columns
`mutate()` adds new columns to the end of your dataset. 

```{r}
starwars_small <- select(starwars, height, mass, birth_year)
head(starwars_small)

starwars_small <- mutate(starwars_small, height_m = height/100) 
mutate(starwars_small, bmi = mass/(height_m^2)) #creates the variable in the dataset

#to only keep the new columns, use transmute()
transmute(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

#using aggregate functions
#if you add NA to anything, the result will be NA, so use na.rm = T to remove it
mutate(starwars_small, tot_mass = sum(mass, na.rm = T))

```


## Summarizing and grouping data
`summarize()` or `summarise()`collapses an entire column of data to a single value. 
```{r}
#creating a new column based on the values of the other column

summarize(starwars, mean_mass = mean(mass, na.rm = T))

#to summarise by certain groups
species_masses <- summarise(group_by(starwars, species), mean_mass = mean(mass, na.rm = T), count = n())
#n will show how many of each species are in there

#to see which species on av weighs the most
arrange(species_masses, desc(mean_mass))


#pipe operator %>%
#function(x , y) is the same as x %>% function(x)
#takes whatever is to the left of the pipe into the other one
#when using dplyr functions, generally always start with dataset

species_masses2 <- starwars %>%
  group_by(species) %>%
  summarize(mean_mass = mean(mass, na.rm = T),
            count = n()) %>%
  arrange(desc(mean_mass))
species_masses2

```


## Sampling a dessignted number of rows
`sample_n()` allows us to sample a random number of rows from our dataset
```{r}
sample_n(starwars, 10)

sample_n(starwars, .2) #for a percentage of rows

nrow(starwars) #total number of rows

```

We can also take a "slice" of our dataset using `slice()` and its related set of functions. Subsetting rows using their positions.
```{r}
starwars %>%
  group_by(gender) %>%
  sample_n(4)

slice(starwars, 50:60) #slice by index value (row number)
slice_head(starwars, n = 10) #first 10 rows
slice_tail(starwars, n = 10) #last 10 rows

slice_min(starwars, order_by = mass, n = 10) #rows with lowest 10 scores
slice_max(starawrs, order_by = mass, n = 10) #rows with highest 10

```

